---
format:
  html:
    theme: cosmo
---

## Objectives {.unnumbered}

- Have a good grasp of `R` code, including:
    - Basic syntax
    - Basic numeric data types (floats, vectors, matrices)
    - Data frames
    - Control flow
    - Functions, including higher-order functions
    - Understanding error messages and debugging
- Be able to write and interpret the output of `R` code covering each of the above.



## Part 1: R Fundamentals

### Tasks

- [x] Print "Hello World" to the console. 
- [ ] Write a function that takes a name, and prints "Hello <name>" to the console.
- [x] Assign the value 42 to a variable called `answer`.
- [x] Create a numeric vector containing the numbers 1 through 5.
- [x] Create a numeric vector containing the numbers 1 through 1000.
- [x] Create a matrix with 10 rows and 10 columns filled with numbers 1 to 100.
- [x] Create a data frame with two columns: one called `name` containing three names, and one called `age` containing three ages.
- [ ] Write a function that returns the square of its input.
- [ ] Use a for loop to print the numbers 1 to 5.
- [ ] Use a for loop to compute the mean of a numeric vector.

```{r}
print("Hello world")


f<-function(name){
  
  print(paste('Hello', name))
}

f('tristan')
```

```{r}
42 <- answer
vector = c(1,2,3,4,5)
vector

vector2 = c(1,1000)
vector2

mat = matrix(1:100, nrow=10, ncol=10)
mat


df <- data.frame(
    name = c("Tristan", "Xiao Bao", "Eason"),
    age = c(20, 24, 31)
)
df


```



## Part 1b: Lists

Recall that a *list* can hold a mixture of types, and that lists turn up all over `R` — in particular, they are how a function returns several results at once, and how we store a collection of records.

- [ ] A function can only return a single object, so when we want to return several results we bundle them into a list. Write a function `summarise_vector(x)` that returns a **named list** with the `mean`, `min`, `max`, and `n` (length) of a numeric vector `x`. Call it on `c(210, 217, 230, 205)` and use `$` to print just the mean from the result.
- [ ] Call `str()` on the value your function returns to confirm its structure, then modify `summarise_vector` so that missing values are ignored (test it on a vector containing `NA`).
- [ ] Below is a list of penguins, each itself a list. Using a `for` loop (or `sapply`), extract the `species` of every penguin into a character vector, and compute the mean flipper length across all of them.

```{r}
penguins <- list(
    list(name = "Flipper", species = "Adelie",  flipper = 210),
    list(name = "Tux",     species = "Gentoo",  flipper = 217),
    list(name = "Pingu",   species = "Emperor", flipper = 230)
)
```

- [ ] The function below is supposed to extract the name of every penguin. Examine its output on the two lists, `penguins` and `penguins2`. Why does it not return the expected values for `penguins2`? Fix the code so it correctly returns penguin names.

```{r}
get_penguin_names <- function(penguins) {
    sapply(penguins, function(penguin) penguin[[1]])
}

penguins2 <- list(
    list(name = "Flipper", species = "Adelie",  flipper = 210),
    list(species = "Gentoo", flipper = 217, name = "Tux"),
    list(flipper = 230, name = "Pingu", species = "Emperor")
)

get_penguin_names(penguins)
get_penguin_names(penguins2)
```


## Part 2: Advanced Array Indexing

In the lecture we saw that we can use *logical indexing* into arrays, i.e.

```{r}
vector <- c(1,2,3,4,5)
elements <- c(TRUE, FALSE, TRUE, TRUE, FALSE)
vector[elements]
```


We can also do *elementwise comparisons*:

```{r}
vector > 3
vector[vector > 3]
```

Consider next this vector:

```{r}
vector <- 1:100
```

Use elementwise indexing to display all elements in the vector that are:

- [ ] $\leq$ 10
- [ ] $>$ 90
- [ ] $\leq 10$ or $> 90$ 
- [ ] $> 10$ and $\leq 90$
- [ ] Odd
- [ ] (Advanced): Prime

Hint: You will need to look up how to *combine* logical indexing statements and also some more *advanced* logical indexing statements.
For the last task look at using `sapply`.

## Part 3: Control Flow

### Part (a): Conditionals

In data science we very often want to make the *same* decision for every element of a vector or column. Writing a for loop with an `if` inside would work, but the *vectorised* conditionals `if_else` and `case_when` are cleaner and faster. Both come from `dplyr`, so load it first:

```{r}
#| message: false
library(dplyr)
```

We will apply them to the following data frame of students and their exam scores:

```{r}
students <- data.frame(
    name = c("Ada", "Alan", "Grace", "Edsger", "Katherine"),
    score = c(72, 65, 43, 38, 91)
)
```

- [ ] Using `if_else`, add a new column `result` to `students` that is `"pass"` when the score is at least 40 and `"fail"` otherwise.
- [ ] Using `case_when`, add a new column `grade` to `students` giving the classification: `"first"` (at least 70), `"2:1"` (at least 60), `"2:2"` (at least 50), `"third"` (at least 40), and `"fail"` otherwise. Remember that conditions are checked in order, so the most specific should come first.
- [ ] Standard `if`/`else` statements make a single decision about a single value. Using `if`, `else if` and `else`, complete the stub function `is_leap_year` below so that it returns `TRUE` if `year` is a *leap year* and `FALSE` otherwise. The rule is:
    - if `year` is divisible by 400, it is a leap year;
    - otherwise, if `year` is divisible by 100, it is *not* a leap year;
    - otherwise, if `year` is divisible by 4, it is a leap year;
    - otherwise, it is *not* a leap year.

::: {.callout-tip}
The modulo operator `%%` gives the remainder of a division, so `year %% 400 == 0` tests whether `year` is divisible by 400.
:::

::: {.callout-note}
## `stopifnot`

`stopifnot()` is a handy way to check your work. Given one or more conditions, it does nothing if they are all `TRUE`, but raises an error as soon as one is `FALSE`. So once your function is correct the chunk below will run silently; until then, the first failing check tells you which case is wrong.
:::

```{r}
#| error: TRUE
is_leap_year <- function(year) {
    FALSE # Replace with correct code.
}

# Self-check: should run without error once is_leap_year is correct
stopifnot(
    is_leap_year(2000) == TRUE,   # divisible by 400
    is_leap_year(2024) == TRUE,   # divisible by 4, not by 100
    is_leap_year(1900) == FALSE,  # divisible by 100, not by 400
    is_leap_year(2100) == FALSE   # divisible by 100, not by 400
)
```

### Part (b): Loops

- [ ] In the next block, write a function that uses a for loop to print a message if an input vector contains NA values. Test it on the five variables supplied below.

```{r}
vec1 <- 1:10
vec2 <- c(1,2,NA,4)
vec3 <- c()
vec4 <- "This is not a vector"
vec5 <- matrix(c(1,2,3,4,NA,6), nrow=2, ncol=3)
```

- [ ] In a new chunk, write a function called `make_square` that, given an integer, will print a square of full stops with side length that integer. Here are some examples:

```r
make_square(3)

# expected output:
...
. .
...

make_square(5)

# expected output:
.....
.   .
.   .
.   .
.....
```

```{r}
make_square <- function(size) {
  return <- ""
  for(row in 1:size) {
    for(col in 1:size) {
      if(row == 1 && col == 1) return <- paste0(return, ".")
      if(row == 1 && col == 2) return <- paste0(return, ".")
      if(row == 1 && col == 3) return <- paste0(return, ".\n")
      if(row == 2 && col == 1) return <- paste0(return, ".")
    }
  }
  return
}

cat(make_square(3))

```



- [ ] (Advanced): Look up *while loops* in `R`. Write a function that uses a while loop to simulate a particle undergoing a *random walk* in 1D. Starting from 0, in each iteration, the particle either takes a step $x \mapsto x + 1$ or $x \mapsto x - 1$, uniformly at random. Terminate the walk when the particle reaches either -10 or +10, and print where it ended up and how many steps it took.
- [ ] (Advanced): Run your function from above multiple times. Display how often the particle ended at -10 and how often it ended at +10. Make sure that your results do not change if the worksheet is run multiple times!

::: {.callout-tip}
Your random walk uses randomly generated numbers, so it will produce different output every time the document is rendered. To make results *reproducible* — the same every time — look up `set.seed()` and call it once at the start of the chunk. We will return to why this matters for reproducibility later in the course.
:::

## Part 4: Pipes

The *pipe* operator `|>` passes the value on its left into the first argument of the function on its right. Chaining pipes lets a sequence of nested function calls be read left-to-right, in the order they actually happen, rather than inside-out.

- [ ] Rewrite the nested expression below using `|>`, so that it reads as a left-to-right pipeline. Check that your piped version gives the same answer. (Note that the last call takes an extra argument — recall that the piped value fills only the *first* argument of each function.)

```{r}
x <- c(4, 9, 16, 25)
round(mean(sqrt(x)), 1)
```

## Part 5: Debugging

In this section we will look at some pathologies of `R` that might require some more work to figure out.
First, try to solve the problems using Google or `R` help files.
Then, feed the problem into an LLM and see whether it comes up with the same solution, a better one (or a worse one!)

### Tasks

- [ ] Modify the below code to return the row sums of only the non-NA values.

```{r}
# A data frame with some missing values
df <- data.frame(a = c(1, 2, 3, NA, 5), b = c(10, NA, 30, 40, 50))

# Compute row-wise sums
row_sums <- apply(df, 1, sum)

print(row_sums)
```

- [ ] Given two vectors `x` and `y` of the same length, the below code prints out the (Euclidean) distance between `x` and `y`, but for some reason it is printing a distance between vectors that are not of the same length. Correct the code.

```{r}
distance <- function(x, y) {
    return(norm(x - y, type="2"))
}
a <- c(1, 2, 3, 4)
b <- c(5, 6)

result <- distance(a, b)
print(result)
```

- [ ] The below code prints out "values are equal" if `a` and `b` are equal, but for some reason it is not working for these example values. Correct the code.
```{r}
a <- 0.1 + 0.2
b <- 0.3

if(a == b) {
    print("Values are equal.")
} else {
    print("Values are not equal.")
}
```

Hint: see the [Assignments and Types](https://soton-fds.github.io/week02.html#assignments-and-types) section of the notes.

- [ ] The below returns the entries of `vec1` for which `vec2 == "y"`. So why does `subset` contain an `NA`? Fix the code.

```{r}
vec1 <- 1:5
vec2 <- c("x", "y", NA, "x", "y")
subset <- vec1[vec2 == "y"]
subset
```